---
title: "SIBYL Round Report"
subtitle: "`r format(Sys.Date(), '%B %Y')`"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
# Run chunks from the project root so tar_load() finds _targets/
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
#| include: false
library(targets)
library(dplyr)
library(tibble)
library(ggplot2)
library(scales)
library(knitr)

`%||%` <- function(a, b) if (is.null(a)) b else a

# Load the four local packages the way _targets.R does, so the qmd works
# whether or not they've been installed for real.
for (pkg in c("packages/judgement", "packages/martin",
              "packages/nowcast",   "packages/sibyldata")) {
  pkg_path <- if (dir.exists(pkg)) pkg else file.path("..", pkg)
  if (dir.exists(pkg_path)) {
    try(pkgload::load_all(pkg_path, quiet = TRUE), silent = TRUE)
  }
}

# Load every required pipeline target the report uses. Each tar_load fails
# fast if the target hasn't been built yet.
tar_load(c(
  round_id, data_source, narrative, horizon,
  handover_forecasts,
  baseline, proposed_adjustments, approved_adjustments,
  projection, projection_description, round_trip_check
))

# Optional targets: load defensively so the report still renders against an
# older pipeline that hasn't built them yet. tar_read_raw() errors if a target
# is absent (or the store is missing), so wrap it to return NULL on failure;
# each call site binds the result explicitly.
soft_tar_load <- function(name) {
  tryCatch(tar_read_raw(name), error = function(e) NULL)
}
# round_metadata: a one-row metadata target the pipeline MAY publish, carrying
# the real coefficient mode / vintage / approval status (Phase 2 contract).
round_metadata        <- soft_tar_load("round_metadata")
# estimation_end: the re-estimation sample end (NULL => frozen coefficients).
estimation_end        <- soft_tar_load("estimation_end")
# database_with_handover: the solve input db; carries a provenance attribute
# when sibyldata built it (live runs). Used for the per-series provenance
# summary. May not survive every db transform, so treat as best-effort.
database_with_handover <- soft_tar_load("database_with_handover")
# Optional stochastic uncertainty bands (opt-in; off the default path).
projection_bands      <- soft_tar_load("projection_bands")
# Optional standalone provenance manifest target (tibble(variable, source_class)).
provenance_manifest   <- soft_tar_load("provenance_manifest")

# Headline aggregates we lead with in every round.
HEADLINE <- c("Y", "RC", "GNE", "LUR", "PTM", "P", "NCR")

# ---------------------------------------------------------------------------
# Derive an honest round-metadata view, preferring an explicit round_metadata
# target when present and degrading gracefully to the targets we already have.
# ---------------------------------------------------------------------------
md_get <- function(field, default = NULL) {
  if (is.null(round_metadata)) return(default)
  # round_metadata may be a one-row tibble/list or a named list.
  if (is.data.frame(round_metadata) && field %in% names(round_metadata)) {
    val <- round_metadata[[field]][1]
  } else if (is.list(round_metadata) && field %in% names(round_metadata)) {
    val <- round_metadata[[field]]
  } else {
    return(default)
  }
  if (is.null(val) || (length(val) == 1L && is.na(val))) default else val
}

# Coefficient mode: honest. The pipeline uses frozen 2019Q3 coefficients only
# when estimation_end is NULL; any non-NULL estimation_end RE-ESTIMATES every
# free coefficient through that quarter (NOT "frozen").
coef_mode <-
  md_get("coefficients",
         if (is.null(estimation_end)) "frozen" else "reestimated")
est_end <-
  md_get("estimation_end",
         if (is.null(estimation_end)) NA_character_ else as.character(estimation_end))

coef_label <- if (identical(coef_mode, "frozen")) {
  "frozen at the model file's 2019Q3 sample (`MARTINMOD_AF.txt`)"
} else {
  paste0("RE-ESTIMATED on data through ",
         if (is.na(est_end)) "an unspecified quarter" else est_end,
         " (free coefficients re-fit on every load; restriction-pinned ",
         "coefficients held)")
}

# Data vintage: a real date/quarter, NOT the scenario label. Prefer an
# explicit vintage from round_metadata; else fall back honestly.
data_vintage <- md_get("data_vintage", NULL)
if (is.null(data_vintage) || is.na(data_vintage) || !nzchar(as.character(data_vintage))) {
  data_vintage <- if (identical(data_source, "fixture")) {
    "unknown (bundled MARTINDATA fixture; no realtime vintage)"
  } else {
    # update_data() stamps a fetch-date "vintage"; we do not have it here
    # unless round_metadata carries it, so be explicit about that.
    "unknown (live fetch; vintage = fetch-date stamp, not a true realtime cut)"
  }
}

# Approval status: was the human gate actually exercised?
approval_status <- md_get("approval_status", NULL)
approval_interactive <- md_get("approval_interactive", NULL)
if (is.null(approval_status)) {
  approval_status <- if (isTRUE(approval_interactive)) {
    "human-approved (interactive review gate)"
  } else if (identical(approval_interactive, FALSE)) {
    "NOT human-reviewed (non-interactive gate; adjustments passed through)"
  } else {
    "unknown (approval gate mode not recorded by the pipeline)"
  }
}

# ---------------------------------------------------------------------------
# Degraded-mode detection. Each flag is honest about WHY the round may not be
# a clean, fully-live, human-approved forecast.
# ---------------------------------------------------------------------------
llm_bypassed <- length(proposed_adjustments) == 0L &&
  !nzchar(Sys.getenv("ANTHROPIC_API_KEY"))
no_adjustments <- length(approved_adjustments) == 0L
fixture_mode   <- identical(data_source, "fixture")

# Per-series provenance: prefer a standalone manifest, else read it off the
# solve-input database via sibyldata::database_provenance().
provenance_tbl <- NULL
if (!is.null(provenance_manifest) && is.data.frame(provenance_manifest)) {
  provenance_tbl <- provenance_manifest
} else if (!is.null(database_with_handover)) {
  provenance_tbl <- tryCatch(
    sibyldata::database_provenance(database_with_handover),
    error = function(e) NULL
  )
}
prov_has <- is.data.frame(provenance_tbl) && nrow(provenance_tbl) > 0L
# A run is "fixture-backfilled" if any series came from the fixture/proxy/dummy
# rather than live data.
fixture_backfilled <- prov_has &&
  any(provenance_tbl$source_class %in%
        c("fixture_fallback", "vendored_wf1", "proxy", "dummy"))

# Convergence of the projection solve (attached by solve_martin()).
proj_conv <- attr(projection, "convergence")
solve_nonconverged <- !is.null(proj_conv) && isFALSE(proj_conv$converged)

degraded <- llm_bypassed || fixture_mode || fixture_backfilled ||
  solve_nonconverged

# Round a numeric to a sensible number of significant figures for headlines.
sig <- function(x, digits = 3) signif(x, digits)
```

```{r degraded-banner}
#| results: asis
if (degraded) {
  cat("::: {.callout-warning title=\"Degraded-mode round\"}\n")
  cat("This round is **not** a clean, fully-live, human-approved forecast. ",
      "Reasons:\n\n", sep = "")
  if (llm_bypassed) {
    cat("- **LLM bypassed.** `ANTHROPIC_API_KEY` is unset, so no narrative ",
        "was translated into add-factors; the round shows the no-adjustment ",
        "baseline.\n", sep = "")
  } else if (no_adjustments) {
    cat("- **No adjustments applied.** The LLM proposed no quantitative ",
        "changes, or all were rejected at review.\n", sep = "")
  }
  if (fixture_mode) {
    cat("- **Fixture data.** `data_source = \"fixture\"`: the round runs on ",
        "the bundled MARTINDATA snapshot, not a live ABS/RBA/FRED fetch.\n",
        sep = "")
  }
  if (fixture_backfilled) {
    cat("- **Fixture/proxy backfill.** Some series were filled from the ",
        "fixture, a vendored `.wf1`, a proxy, or a dummy rather than live ",
        "data (see the provenance summary below).\n", sep = "")
  }
  if (solve_nonconverged) {
    cat("- **Solve did not fully converge.** The projection carries ",
        proj_conv$n_nonfinite %||% NA, " non-finite cell(s) inside the ",
        "solved window; treat the affected paths with caution.\n", sep = "")
  }
  cat(":::\n\n")
}
```

# Narrative

```{r narrative}
#| results: asis
cat("> ", narrative, "\n", sep = "")
```

# Round configuration

- **Round ID:** `r round_id`
- **Data source:** `r data_source` `r if (identical(data_source, "fixture")) "(bundled MARTINDATA snapshot)"`
- **Horizon:** `r horizon[1]` through `r horizon[2]`
- **Coefficients:** `r coef_label`
- **Approval status:** `r approval_status`
- **Adjustments proposed:** `r length(proposed_adjustments)`
- **Adjustments approved:** `r length(approved_adjustments)`

# Proposed and approved adjustments

```{r adjustments}
if (length(approved_adjustments) == 0L) {
  cat("_No adjustments applied this round._",
      if (length(proposed_adjustments) == 0L) {
        " The LLM was either bypassed (`ANTHROPIC_API_KEY` unset) or proposed no quantitative changes from the narrative."
      } else "",
      "\n")
} else {
  approved_tbl <- as_tibble_adjustments(approved_adjustments) |>
    select(equation, quarter, value, rationale, tail, confidence) |>
    arrange(equation, quarter)
  knitr::kable(approved_tbl, digits = 3)
}
```

```{r exogenize}
#| results: asis
# Show the exogenize list (variables pinned to the baseline path) so the human
# can see what was held fixed. The attribute survives the approval gate.
exo <- attr(approved_adjustments, "exogenize") %||% character(0)
if (length(exo) > 0L) {
  cat("**Variables exogenised (pinned to the baseline path):** ",
      paste(exo, collapse = ", "), "\n", sep = "")
}
```

# Headline aggregates: baseline vs scenario

```{r headlines-plot}
#| fig-width: 9
#| fig-height: 6
plot_df <- dplyr::bind_rows(
  baseline   |> mutate(series = "baseline"),
  projection |> mutate(series = "with_adjustments")
) |>
  filter(variable %in% HEADLINE) |>
  mutate(
    year    = as.integer(substr(quarter, 1, 4)),
    qnum    = as.integer(substr(quarter, 6, 6)),
    qdate   = as.Date(sprintf("%04d-%02d-15", year, (qnum - 1) * 3 + 2))
  )

if (length(approved_adjustments) == 0L) {
  plot_df <- plot_df |> filter(series == "baseline")
}

p <- ggplot(plot_df, aes(qdate, value, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_colour_manual(values = c(baseline = "grey50",
                                  with_adjustments = "steelblue")) +
  labs(x = NULL, y = NULL, colour = NULL,
       title = "Solved projections, headline aggregates",
       subtitle = paste("Horizon", horizon[1], "to", horizon[2])) +
  theme_minimal() +
  theme(legend.position = "bottom")

# If a stochastic uncertainty band target exists, overlay it (central + ribbon).
band_overlaid <- FALSE
if (!is.null(projection_bands) && is.data.frame(projection_bands) &&
    all(c("variable", "quarter", "lower", "upper") %in% names(projection_bands))) {
  band_df <- projection_bands |>
    filter(variable %in% HEADLINE) |>
    mutate(
      year  = as.integer(substr(quarter, 1, 4)),
      qnum  = as.integer(substr(quarter, 6, 6)),
      qdate = as.Date(sprintf("%04d-%02d-15", year, (qnum - 1) * 3 + 2))
    )
  if (nrow(band_df) > 0L) {
    p <- p +
      geom_ribbon(
        data = band_df,
        aes(qdate, ymin = lower, ymax = upper),
        inherit.aes = FALSE, fill = "steelblue", alpha = 0.15
      )
    band_overlaid <- TRUE
  }
}
p
```

```{r headline-table}
#| results: asis
if (length(approved_adjustments) > 0L) {
  # Select the horizon-end quarter EXPLICITLY (the latest quarter common to
  # both paths) rather than trusting row order: last(diff) silently returns
  # whatever happens to be the final row of the join.
  joined <- inner_join(
    baseline   |> select(variable, quarter, baseline = value),
    projection |> select(variable, quarter, with_adjustments = value),
    by = c("variable", "quarter")
  ) |>
    filter(variable %in% HEADLINE) |>
    mutate(diff = with_adjustments - baseline)

  end_q <- max(joined$quarter)  # "yyyyQq" sorts lexically within a century

  diff_df <- joined |>
    group_by(variable) |>
    summarise(
      max_abs_diff = max(abs(diff), na.rm = TRUE),
      diff_at_end  = diff[quarter == end_q][1],
      .groups = "drop"
    ) |>
    # Round headline diffs to sensible significant figures.
    mutate(
      max_abs_diff = sig(max_abs_diff, 3),
      diff_at_end  = sig(diff_at_end, 3)
    )

  cat("## Diffs vs baseline at end of horizon (", end_q, ")\n\n", sep = "")
  print(knitr::kable(diff_df, digits = 3))
  cat("\n_The horizon-end column (", end_q, ") is the deepest extrapolation ",
      "point in the run — furthest from the last quarter of hard data and so ",
      "the least certain figure in the table._\n", sep = "")
} else {
  cat("_No diff to show — no adjustments applied._\n")
}
```

# Uncertainty

```{r uncertainty-note}
#| results: asis
cat("SIBYL produces a **single model-consistent path**. It does **not** ",
    "quantify parameter uncertainty, trend/equilibrium uncertainty, or data ",
    "revisions — those are out of scope for this layer. The figures above are ",
    "point projections conditional on the chosen coefficients (",
    coef_mode, "), the exogenous-variable assumptions, and the approved ",
    "add-factors.\n\n", sep = "")
if (band_overlaid) {
  bm <- attr(projection_bands, "band_method") %||% "unknown"
  nd <- attr(projection_bands, "n_draws") %||% NA
  cat("A stochastic band is overlaid on the headline chart (shaded ribbon, ",
      "2.5%/97.5% edges) from `martin::solve_martin_stochastic()` — method: **",
      bm, "**, draws: **", nd, "**. This captures equation-disturbance / ",
      "coefficient sampling only; it is still conditional on the exogenous ",
      "and add-factor assumptions.\n", sep = "")
} else {
  cat("_No stochastic uncertainty band was computed for this round (opt-in via ",
      "`martin::solve_martin_stochastic()`)._\n", sep = "")
}
```

# Nowcast handover summary

```{r handover}
if (nrow(handover_forecasts) == 0L) {
  cat("_No handover forecasts produced (database had no ragged edge)._\n")
} else {
  knitr::kable(
    handover_forecasts |>
      filter(variable %in% HEADLINE) |>
      mutate(quarter = as.character(quarter)) |>
      arrange(variable, quarter),
    digits = 3
  )
}
```

# LLM description of the projection

```{r description}
#| results: asis
cat(projection_description, "\n")
```

# Round-trip check

```{r round-trip}
if (nrow(round_trip_check) == 0L) {
  cat("_No round-trip check performed (no adjustments applied, or LLM bypassed)._\n")
} else {
  overall <- attr(round_trip_check, "overall_match")
  if (!is.null(overall)) {
    cat("**Overall match:** _", overall, "_\n\n", sep = "")
  }
  knitr::kable(round_trip_check)
}
```

## Mechanical fidelity audit

This is the **LLM-independent** fidelity check: for every adjustment that
declared a target variable and an expected direction, it compares that
declared direction against the realised projection-minus-baseline diff at the
horizon end. It does not rely on any model-written prose, so it cannot be
talked around by a persuasive description.

```{r mechanical-audit}
#| results: asis
mech <- tryCatch(
  judgement::mechanical_audit(approved_adjustments, projection, baseline),
  error = function(e) NULL
)
if (is.null(mech) || nrow(mech) == 0L) {
  cat("_No mechanical audit rows — no adjustment declared a target ",
      "variable + expected direction this round._\n", sep = "")
} else {
  mech_show <- mech |>
    mutate(realised_diff = sig(realised_diff, 3))
  print(knitr::kable(mech_show, digits = 3))
  n_disagree <- sum(mech$agrees %in% FALSE)
  if (n_disagree > 0L) {
    cat("\n**", n_disagree, " adjustment(s) realised the WRONG direction ",
        "vs their declared target** — a translation or magnitude failure the ",
        "refinement loop or a human should reconcile.\n", sep = "")
  }
}
```

## Narrative coherence diagnostics

Where the audit returns `disagree`, this section attempts to tell apart
two distinct failure modes:

* **translation_gap** — the narrative quantified a target on variable X
  and the projection missed it. Likely indicates an AF magnitude or
  equation-choice issue that the refinement loop should address (or
  that needs human revision).
* **model_response** — the narrative asserted no change to variable X
  but the projection moved it anyway. If X is endogenous (e.g. NCR via
  the Taylor Rule when LUR shifts) this is MARTIN behaving as designed,
  not a translation failure. Add a cancelling AF on X if you really
  need to suppress the response.

`not_addressed` claims — ones the blind describer never restated — are also
surfaced here so a silently-dropped narrative point is visible (matching the
dashboard) rather than quietly filtered out.

```{r diagnose}
#| results: asis
if (nrow(round_trip_check) == 0L) {
  cat("_Nothing to diagnose._\n")
} else {
  diag <- judgement::diagnose_audit(
    audit      = round_trip_check,
    projection = projection,
    baseline   = baseline
  )
  # Surface rows that need attention. `not_addressed` is included so a claim
  # the describer silently dropped is visible rather than filtered away.
  attention <- diag[diag$category %in%
                     c("translation_gap", "model_response",
                       "narrative_conflict", "unclassified",
                       "not_addressed"), ,
                    drop = FALSE]
  if (nrow(attention) == 0L) {
    cat("_All audit claims accepted; no diagnostics needed._\n")
  } else {
    show <- attention[, c("claim", "category", "variable",
                          "diff_at_end", "explanation"), drop = FALSE]
    show$diff_at_end <- sig(show$diff_at_end, 3)
    print(knitr::kable(show, digits = 3))
  }
}
```

# Provenance

```{r provenance}
#| results: asis
cat("- **Generated:** ", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n", sep = "")
cat("- **Coefficients:** ", coef_label, "\n", sep = "")
cat("- **Data source:** ", data_source, "\n", sep = "")
cat("- **Data vintage:** ", as.character(data_vintage), "\n", sep = "")
cat("- **Approval status:** ", approval_status, "\n", sep = "")
sc <- attr(projection, "scenario")
if (!is.null(sc)) {
  cat("- **Scenario label:** ", sc,
      " (a run label, not a data vintage)\n", sep = "")
}
if (solve_nonconverged) {
  cat("- **Solve convergence:** NON-CONVERGED (",
      proj_conv$n_nonfinite %||% NA, " non-finite cell(s))\n", sep = "")
} else if (!is.null(proj_conv)) {
  cat("- **Solve convergence:** converged\n")
}
```

```{r provenance-series}
#| results: asis
# Honest per-series provenance summary (live vs fixture vs proxy vs dummy)
# from the provenance manifest when available.
if (prov_has) {
  cat("## Per-series provenance\n\n")
  prov_summary <- provenance_tbl |>
    count(source_class, name = "n_series") |>
    arrange(desc(n_series))
  print(knitr::kable(prov_summary))
  cat("\n_`live` = fetched from a public source this vintage; ",
      "`fixture_fallback` = backfilled from the bundled MARTINDATA snapshot; ",
      "`vendored_wf1` = spliced from a vendored EViews workfile; ",
      "`proxy` = a stand-in series; `dummy` = a deterministic 0/1 indicator; ",
      "`derived` = computed from other series; `unknown` = off-catalogue._\n",
      sep = "")
} else {
  cat("## Per-series provenance\n\n")
  cat("_No provenance manifest available for this round (the solve-input ",
      "database carried no `provenance` attribute — e.g. a hand-built or ",
      "fixture-only database)._\n", sep = "")
}
```

> This report is generated by the pipeline in `_targets.R`. The LLM is a
> translator and drafter only — every numerical adjustment is meant to be
> reviewed and approved by a human (via `judgement::review_and_approve()`)
> before MARTIN sees it. The **Approval status** above states whether that
> gate was actually exercised this run; a non-interactive gate passes
> adjustments through unreviewed and is flagged in the degraded-mode banner.
