---
title: "SIBYL Round Report"
subtitle: "`r format(Sys.Date(), '%B %Y')`"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
# Run chunks from the project root so tar_load() finds _targets/
knitr:
  opts_knit:
    root.dir: ".."
---

```{r setup}
#| include: false
library(targets)
library(dplyr)
library(tibble)
library(ggplot2)
library(scales)
library(knitr)

# Load the four local packages the way _targets.R does, so the qmd works
# whether or not they've been installed for real.
for (pkg in c("packages/judgement", "packages/martin",
              "packages/nowcast",   "packages/sibyldata")) {
  pkg_path <- if (dir.exists(pkg)) pkg else file.path("..", pkg)
  if (dir.exists(pkg_path)) {
    try(pkgload::load_all(pkg_path, quiet = TRUE), silent = TRUE)
  }
}

# Load every pipeline target the report uses. Each tar_load fails fast if
# the target hasn't been built yet.
tar_load(c(
  round_id, data_source, narrative, horizon,
  handover_forecasts,
  baseline, proposed_adjustments, approved_adjustments,
  projection, projection_description, round_trip_check
))

# Headline aggregates we lead with in every round.
HEADLINE <- c("Y", "RC", "GNE", "LUR", "PTM", "P", "NCR")
```

# Narrative

```{r narrative}
#| results: asis
cat("> ", narrative, "\n", sep = "")
```

# Round configuration

- **Round ID:** `r round_id`
- **Data source:** `r data_source` `r if (data_source == "fixture") "(bundled MARTINDATA snapshot)"`
- **Horizon:** `r horizon[1]` through `r horizon[2]`
- **Adjustments proposed:** `r length(proposed_adjustments)`
- **Adjustments approved:** `r length(approved_adjustments)`

# Proposed and approved adjustments

```{r adjustments}
if (length(approved_adjustments) == 0L) {
  cat("_No adjustments applied this round._",
      if (length(proposed_adjustments) == 0L) {
        " The LLM was either bypassed (`ANTHROPIC_API_KEY` unset) or proposed no quantitative changes from the narrative."
      } else "",
      "\n")
} else {
  approved_tbl <- as_tibble_adjustments(approved_adjustments) |>
    select(equation, quarter, value, rationale, tail, confidence) |>
    arrange(equation, quarter)
  knitr::kable(approved_tbl, digits = 3)
}
```

# Headline aggregates: baseline vs scenario

```{r headlines-plot}
#| fig-width: 9
#| fig-height: 6
plot_df <- dplyr::bind_rows(
  baseline   |> mutate(series = "baseline"),
  projection |> mutate(series = "with_adjustments")
) |>
  filter(variable %in% HEADLINE) |>
  mutate(
    year    = as.integer(substr(quarter, 1, 4)),
    qnum    = as.integer(substr(quarter, 6, 6)),
    qdate   = as.Date(sprintf("%04d-%02d-15", year, (qnum - 1) * 3 + 2))
  )

if (length(approved_adjustments) == 0L) {
  plot_df <- plot_df |> filter(series == "baseline")
}

ggplot(plot_df, aes(qdate, value, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_colour_manual(values = c(baseline = "grey50",
                                  with_adjustments = "steelblue")) +
  labs(x = NULL, y = NULL, colour = NULL,
       title = "Solved projections, headline aggregates",
       subtitle = paste("Horizon", horizon[1], "to", horizon[2])) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r headline-table}
#| results: asis
if (length(approved_adjustments) > 0L) {
  diff_df <- inner_join(
    baseline   |> select(variable, quarter, baseline = value),
    projection |> select(variable, quarter, with_adjustments = value),
    by = c("variable", "quarter")
  ) |>
    filter(variable %in% HEADLINE) |>
    mutate(diff = with_adjustments - baseline) |>
    group_by(variable) |>
    summarise(
      max_abs_diff = max(abs(diff), na.rm = TRUE),
      diff_at_end  = last(diff),
      .groups = "drop"
    )
  cat("## Diffs vs baseline at end of horizon\n\n")
  print(knitr::kable(diff_df, digits = 3))
} else {
  cat("_No diff to show — no adjustments applied._\n")
}
```

# Nowcast handover summary

```{r handover}
if (nrow(handover_forecasts) == 0L) {
  cat("_No handover forecasts produced (database had no ragged edge)._\n")
} else {
  knitr::kable(
    handover_forecasts |>
      filter(variable %in% HEADLINE) |>
      mutate(quarter = as.character(quarter)) |>
      arrange(variable, quarter),
    digits = 3
  )
}
```

# LLM description of the projection

```{r description}
#| results: asis
cat(projection_description, "\n")
```

# Round-trip check

```{r round-trip}
if (nrow(round_trip_check) == 0L) {
  cat("_No round-trip check performed (no adjustments applied, or LLM bypassed)._\n")
} else {
  overall <- attr(round_trip_check, "overall_match")
  if (!is.null(overall)) {
    cat("**Overall match:** _", overall, "_\n\n", sep = "")
  }
  knitr::kable(round_trip_check)
}
```

## Narrative coherence diagnostics

Where the audit returns `disagree`, this section attempts to tell apart
two distinct failure modes:

* **translation_gap** — the narrative quantified a target on variable X
  and the projection missed it. Likely indicates an AF magnitude or
  equation-choice issue that the refinement loop should address (or
  that needs human revision).
* **model_response** — the narrative asserted no change to variable X
  but the projection moved it anyway. If X is endogenous (e.g. NCR via
  the Taylor Rule when LUR shifts) this is MARTIN behaving as designed,
  not a translation failure. Add a cancelling AF on X if you really
  need to suppress the response.

```{r diagnose}
#| results: asis
if (nrow(round_trip_check) == 0L) {
  cat("_Nothing to diagnose._\n")
} else {
  diag <- judgement::diagnose_audit(
    audit      = round_trip_check,
    projection = projection,
    baseline   = baseline
  )
  # Surface only rows that need attention.
  attention <- diag[diag$category %in%
                     c("translation_gap", "model_response",
                       "narrative_conflict", "unclassified"), ,
                    drop = FALSE]
  if (nrow(attention) == 0L) {
    cat("_All audit claims accepted or addressed; no diagnostics needed._\n")
  } else {
    show <- attention[, c("claim", "category", "variable",
                          "diff_at_end", "explanation"), drop = FALSE]
    print(knitr::kable(show, digits = 3))
  }
}
```

# Provenance

```{r provenance}
#| results: asis
cat("- **Generated:** ", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n", sep = "")
cat("- **Coefficients:** frozen (`MARTINMOD_AF.txt`)\n")
cat("- **Database vintage:** ", attr(baseline, "scenario"), "\n", sep = "")
```

> This report is generated by the pipeline in `_targets.R`. The LLM is a
> translator and drafter only — every numerical adjustment is reviewed and
> approved by a human (via `judgement::review_and_approve()`) before MARTIN
> sees it. In this run the human-approval step is non-interactive; flip the
> `interactive` arg in `_targets.R` to gate manually.
